Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-06-28 article-models-ttinternal.tex
fixed broken references
small improvements; starting to write the rules of weakening algebras

\begin{proof}
We only prove the first inference rule. Let $\jfam{{\Gamma}{A}}{Q}$ be a family.
Then we have the judgmental equalities
\begin{align*}
\jcomp{}{e_0}{\jcomp{}{e_0}{Q}}
& \jdeq
  \jcomp{}{\idtm{\jcomp{}{e_0}{P}}}{\jcomp{}{e_0}{\jcomp{}{e_0}{Q}}}
  \tag{by \autoref{idfunc-precomp}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{e_0}{\idtm{\jcomp{}{e_0}{P}}}}{\jcomp{}{e_0}{Q}}
  \tag{by \autoref{lem:composition-threesome}}
  \\
& \jdeq
  \jcomp{}{\jcomp{}{\jvcomp{}{e_0}{\idtm{\jcomp{}{e_0}{P}}}}{e_0}}{Q}
  \tag{by \autoref{lem:jcomp-jcomp}}
  \\
& \jdeq
  \jcomp{}{\jcomp{}{\jvcomp{}{\idtm{A}}{e_1}}{e_0}}{Q}
  \tag{by \autoref{eq:extalg-eq1}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{\idtm{A}}{e_1}}{\jcomp{}{e_0}{Q}}
  \tag{by \autoref{lem:jcomp-jcomp}}
  \\
& \jdeq
  \jcomp{}{e_1}{\jcomp{}{\idtm{A}}{\jcomp{}{e_0}{Q}}}
  \tag{by \autoref{lem:composition-threesome}}
  \\
& \jdeq
  \jcomp{}{e_1}{\jcomp{}{e_0}{Q}}.
  \tag{by \autoref{idfunc-precomp}}
\end{align*}
\end{proof}

@@

More precisely, for $x_0:\cftalgc{\cftalg{A}}$, 
$x_1:\subst{x_0}{\cftalgf{\cftalg{A}}}$ and 
$x_2:\subst{{x_1}{{x_0}{\cftctxext[\extalg{A}]}}}{\cftalgf{\cftalg{A}}}$ 
we have the judgmental equality
\begin{align*}
\subst{{x_2}{{x_1}{\cftfamext[\extalg{A}]}}}{{x_0}{\cftfamext[\extalg{A}]}}
& \jdeq
  \subst
    {x_2}
    {{{x_1}{{x_0}{\cftfamext[\extalg{A}]}}}{\cftfamext[\extalg{A}]}}.
  \intertext{%
and for $y_0:\cftalgf{\extalg{A}}$, 
$y_1:\subst{y_0}{\jcomp{}{\cftfamext[\extalg{A}]}{\cftalgf{\extalg{A}}}}$ and 
$y_2:
  \subst
    {y_1}
    { {y_0}
      { \jcomp
          {}
          {\cftfamext[\extalg{A}]}
          {\jcomp{}{\cftfamext[\extalg{A}]}{\cftalgf{\extalg{A}}}}
        }
      }
$
we have the judgmental equality}
\subst
  {{y_2}{{y_1}{\jcomp{}{\cftfamext[\extalg{A}]}{\cftfamext[\extalg{A}]}}}}
  {{y_0}{\cftfamext[\extalg{A}]}}
& \jdeq
  \subst{y_2}{{{y_1}{{y_0}{\cftfamext[\extalg{A}]}}}{\cftfamext[\extalg{A}]}}.
\end{align*}
\end{lem}

@@

\begin{defn}
A quintuple $(\cftalgc{\cftalg{A}},\cftalgf{\cftalg{A}},\cftalgt{\cftalg{A}},
\cftempc{\cftalg{A}},\cftempf{\cftalg{A}})$ consisting of
\begin{enumerate}
\item A family $\cftalgc{\cftalg{A}}$ in context $\Gamma$,
\item A family $\cftalgf{\cftalg{A}}$ in context $\ctxext{\Gamma}{\cftalgc{\cftalg{A}}}$,
\item A family $\cftalgt{\cftalg{A}}$ in context $\ctxext{{\Gamma}{A}}{\cftalgf{\cftalg{A}}}$,
\item A term $\cftempc{\cftalg{A}}$ of $\cftalgc{\cftalg{A}}$ in context $\Gamma$,
\item A term $\cftempf{\cftalg{A}}$ of $\cftalgf{\cftalg{A}}$ in context $\ctxext{\Gamma}{\cftalgc{\cftalg{A}}}$,
\end{enumerate}
is said to be an \emph{empty-object algebra in context $\Gamma$} if additionally
the following judgmental equalities hold:
\begin{align}
\jfameq*{\Gamma}{\subst{\cftempc{\cftalg{A}}}{\cftalgf{\cftalg{A}}}}{\cftalgc{\cftalg{A}}}
  \label{empalg-eq1}
  \\
\jtermeq*{\Gamma}{\cftalgc{\cftalg{A}}}{\subst{\cftempc{\cftalg{A}}}{\cftempf{\cftalg{A}}}}{\cftempc{\cftalg{A}}}.
  \label{empalg-eq2}
  \\
\jfameq*{{\Gamma}{\cftalgc{\cftalg{A}}}}{\subst{\cftempf{\cftalg{A}}}{\cftalgt{\cftalg{A}}}}{\emptyf}
\end{align}
When $(\cftalgc{\cftalg{A}},\cftalgf{\cftalg{A}},\cftalgt{\cftalg{A}},
\cftempc{\cftalg{A}},\cftempf{\cftalg{A}})$ is an empty-object algebra
in context $\Gamma$, we will
also refer to the pair $(\cftempc{\cftalg{A}},\cftempf{\cftalg{A}})$ as the
\emph{empty object of $\cftalg{A}$}.
\end{defn}

@@

\subsection{Weakening term-algebras}
Weakening term-algebras will be term-algebras with certain added structure.
Although strictly speaking one could formulate a notion of weakening algebra
which only depends on extension algebras and which omits both empty objects
and terms, we shall not do so here.

Let $(A,P,T,\epsilon_0,\epsilon_1,\phi_0,\phi_1)$ be a term-algebra in context
$\Gamma$. The weakening homomorphism $\omega$ will be a term-homomorphism from 
$\ctxwk{P}{\mathcal{F}_\mathcal{A}}$ to 
$\jcomp{}{\epsilon_0}{\mathcal{F}_\mathcal{A}}$ satisfying certain extra
judgmental equalities. The goal of this subsection is to spell these out.

@@

\subsubsection{The definition of weakening algebras}

>>>>

\begin{proof}
We only prove the first inference rule. Let $\jfam{{\Gamma}{A}}{Q}$ be a family.
Then we have the judgmental equalities
\begin{align*}
\jcomp{}{e_0}{\jcomp{}{e_0}{Q}}
& \jdeq
  \jcomp{}{\idtm{\jcomp{}{e_0}{P}}}{\jcomp{}{e_0}{\jcomp{}{e_0}{Q}}}
  \tag{by \autoref{precomp-idtm-f}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{e_0}{\idtm{\jcomp{}{e_0}{P}}}}{\jcomp{}{e_0}{Q}}
  \tag{by \autoref{lem:composition-threesome}}
  \\
& \jdeq
  \jcomp{}{\jcomp{}{\jvcomp{}{e_0}{\idtm{\jcomp{}{e_0}{P}}}}{e_0}}{Q}
  \tag{by \autoref{lem:jcomp-jcomp}}
  \\
& \jdeq
  \jcomp{}{\jcomp{}{\jvcomp{}{\idtm{A}}{e_1}}{e_0}}{Q}
  \tag{by \autoref{eq:extalg-eq1}}
  \\
& \jdeq
  \jcomp{}{\jvcomp{}{\idtm{A}}{e_1}}{\jcomp{}{e_0}{Q}}
  \tag{by \autoref{lem:jcomp-jcomp}}
  \\
& \jdeq
  \jcomp{}{e_1}{\jcomp{}{\idtm{A}}{\jcomp{}{e_0}{Q}}}
  \tag{by \autoref{lem:composition-threesome}}
  \\
& \jdeq
  \jcomp{}{e_1}{\jcomp{}{e_0}{Q}}.
  \tag{by \autoref{precomp-idtm-f}}
\end{align*}
\end{proof}

@@

More precisely, for $x_0:\cftalgc{\cftalg{A}}$, 
$x_1:\subst{x_0}{\cftalgf{\cftalg{A}}}$ and 
$x_2:\subst{{x_1}{{x_0}{\cftctxext[\extalg{A}]}}}{\cftalgf{\cftalg{A}}}$ 
we have the judgmental equality
\begin{align*}
\subst{{x_2}{{x_1}{\cftfamext[\extalg{A}]}}}{{x_0}{\cftctxext[\extalg{A}]}}
& \jdeq
  \subst
    {x_2}
    {{{x_1}{{x_0}{\cftctxext[\extalg{A}]}}}{\cftctxext[\extalg{A}]}}.
  \intertext{%
and for $y_0:\cftalgf{\extalg{A}}$, 
$y_1:\subst{y_0}{\jcomp{}{\cftfamext[\extalg{A}]}{\cftalgf{\extalg{A}}}}$ and 
$y_2:
  \subst
    {y_1}
    { {y_0}
      { \jcomp
          {}
          {\cftfamext[\extalg{A}]}
          {\jcomp{}{\cftfamext[\extalg{A}]}{\cftalgf{\extalg{A}}}}
        }
      }
$
we have the judgmental equality}
\subst
  {{y_2}{{y_1}{\jcomp{}{\cftfamext[\extalg{A}]}{\cftfamext[\extalg{A}]}}}}
  {{y_0}{\cftfamext[\extalg{A}]}}
& \jdeq
  \subst{y_2}{{{y_1}{{y_0}{\cftfamext[\extalg{A}]}}}{\cftfamext[\extalg{A}]}}.
\end{align*}
\end{lem}

@@

\begin{defn}
A quintuple $(\cftalgc{\cftalg{A}},\cftalgf{\cftalg{A}},\cftalgt{\cftalg{A}},
\cftempc{\cftalg{A}},\cftempf{\cftalg{A}})$ consisting of
\begin{enumerate}
\item A family $\cftalgc{\cftalg{A}}$ in context $\Gamma$,
\item A family $\cftalgf{\cftalg{A}}$ in context $\ctxext{\Gamma}{\cftalgc{\cftalg{A}}}$,
\item A family $\cftalgt{\cftalg{A}}$ in context $\ctxext{{\Gamma}{A}}{\cftalgf{\cftalg{A}}}$,
\item A term $\cftempc{\cftalg{A}}$ of $\cftalgc{\cftalg{A}}$ in context $\Gamma$,
\item A term $\cftempf{\cftalg{A}}$ of $\cftalgf{\cftalg{A}}$ in context $\ctxext{\Gamma}{\cftalgc{\cftalg{A}}}$,
\end{enumerate}
is said to be an \emph{empty-object algebra in context $\Gamma$} if additionally
the following judgmental equalities hold:
\begin{align}
\jfameq*{\Gamma}{\subst{\cftempc{\cftalg{A}}}{\cftalgf{\cftalg{A}}}}{\cftalgc{\cftalg{A}}}
  \label{empalg-eq1}
  \\
\jtermeq*{\Gamma}{\cftalgc{\cftalg{A}}}{\subst{\cftempc{\cftalg{A}}}{\cftempf{\cftalg{A}}}}{\cftempc{\cftalg{A}}}.
  \label{empalg-eq2}
%  \\
%\jfameq*{{\Gamma}{\cftalgc{\cftalg{A}}}}{\subst{\cftempf{\cftalg{A}}}{\cftalgt{\cftalg{A}}}}{\emptyf}
\end{align}
When $(\cftalgc{\cftalg{A}},\cftalgf{\cftalg{A}},\cftalgt{\cftalg{A}},
\cftempc{\cftalg{A}},\cftempf{\cftalg{A}})$ is an empty-object algebra
in context $\Gamma$, we will
also refer to the pair $(\cftempc{\cftalg{A}},\cftempf{\cftalg{A}})$ as the
\emph{empty object of $\cftalg{A}$}.
\end{defn}

@@

\subsection{Weakening algebras}
Weakening algebras will be term-algebras with certain added structure.
Although strictly speaking one could formulate a notion of weakening algebra
which only depends on extension algebras and which omits both empty objects
and terms, we shall not do so here.

Let $\cftalg{A}$ be a term-algebra in context $\Gamma$. Weakening will be a
homomorphism
\begin{equation*}
\jhom
  {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
  {\ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_A}}}
  {\jcomp{}{\cftctxext}{\cftalg{F_A}}}
  {\cftwk{\cftalg{A}}}
\end{equation*}
satisfying judgmental equalities which express the compatibility with extension,
the empty object and itself. Thus, the rules we interpret for weakening are
the rules in sections \autoref{comp-we,comp-w0,comp-ww,comp-ew,comp-0w}. The
definition of weakening algebras can be found in \autoref{sec:cftwkalg-defn}.

\subsubsection{Weakening preserves itself}
The following diagram displays the compatibility of weakening with itself:
\begin{equation*}
\begin{tikzcd}[column sep=huge]
\ctxwk{\cftalgf{\cftalg{A}}}{{\cftalgf{\cftalg{F_A}}}{\cftalg{F_{F_A}}}}
  \ar{d}
    [swap]{ \ctxwk
        {{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}
        {\boldsymbol{\mathcal{F}}_{\cftwk{\cftalg{A}}}}
      }
  \ar{r}{\ctxwk{\cftalgf{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}}
& \ctxwk{\cftalgf{\cftalg{A}}}{\cftalg{F_{F_{F_A}}}}
  \ar{d}{\jcomp{}{\cftctxext}{\boldsymbol{\mathcal{F}}_{\cftwk{\cftalg{A}}}}}
  \\
\jcomp{}{\cftwkc{A}}{
\ctxwk{{\cftalgf{\cftalg{A}}}{\cftalgf{\cftalg{F_A}}}}{\cftalg{F_{F_{F_A}}}}}
  \ar{r}[swap]{\jcomp{}{\cftwkc{A}}{\jcomp{}{\cftctxext}{\cftwk{A}}}}
& \jcomp{}{\cftwkc{A}}{\cftalg{F_{F_{F_{F_A}}}}}
\end{tikzcd}
\end{equation*}

\subsubsection{Currying for weakening}
\begin{equation*}
\jhomeq
  { {{{\Gamma}{\cftalgc{\cftalg{A}}}}{\cftalgf{\cftalg{A}}}}
    {\cftalgf{\cftalg{F_A}}}
    }
  { \ctxwk
      {\cftalgf{\cftalg{F_A}}}
      {\cftalg{F_{F_A}}}
    }
  { \cftalg{F_{F_{F_A}}}
    }
  { \jcomp{}{\cftfamext}{\cftwk{\cftalg{A}}}
    }
  { \jcomp{}{\cftwk{\cftalg{A}}}{\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}}
    }
\end{equation*}

\subsubsection{Weakening by the empty family}
\begin{equation*}
\jhomeq
  { {\Gamma}{\cftalgc{\cftalg{A}}}
    }
  { \cftalg{F_A}
    }
  { \cftalg{F_A}
    }
  { \subst{\cftempf{\cftalg{A}}}{\cftwk{\cftalg{A}}}
    }
  { id_{\cftalg{F_A}}
    }
\end{equation*}

\subsubsection{The definition of weakening algebras}\label{sec:cftwkalg-defn}

\subsubsection{Derivable properties of weakening algebras}
We should theoremize the following statements:
\begin{enumerate}
\item If $\cftalg{A}$ is a weakening algebra, then so is $\cftalg{F_A}$ if we take
$\jcomp{}{\cftctxext}{\cftwk{\cftalg{A}}}$ as its weakening.
\end{enumerate}

Let $\cftalg{A}$ be a term-algebra in context
$\Gamma$. The weakening homomorphism $\omega$ will be a term-homomorphism from 
$\ctxwk{P}{\mathcal{F}_\mathcal{A}}$ to 
$\jcomp{}{\epsilon_0}{\mathcal{F}_\mathcal{A}}$ satisfying certain extra
judgmental equalities. The goal of this subsection is to spell these out.

@@
